Renewal and stability in populations structured by remaining years of life

### 2014-09-12 ThanoRepro/ThanoRepro.tex
started to peck at empirical findings. Maybe it'd be cooler to get into eigenvalues more...

\section*{Some preliminary empirical findings}
The thanatological renewal equation and the discrete projection matrix can be
put to work with data. At this time we are only ready to report some early
results, and we do not yet have explanations some of the patterns that we
report. Of our 1834 population-years, we optimize $r$ from both
\eqref{eq:thanoren} and \eqref{eq:lotka}\footnote{It is possible to optimize $r$ using a variety of approximations, or by using a generic optimizer. We have used the method proposed by \citet{coale1957new} for Lotka's $r$ and a modified version of the same for
the thanatological $r$. Details and / or implementation code available on
request.}. In this sample, both versions of $r$ are only plausibly equal in a
single instance. Usually thantological $r$ is greater than Lotka's $r$ (1373
cases). When Lotka's $r$ is positive (693 cases), thanatological $r$ is greater
just over of 50\% of the time (356), but when Lotka's $r$ is negative,
thanatological $r$ is the greater of the two around 90\% of the time (1017
cases). These two approximations of $r$ are of opposite sign 138 times. We
provide a comparison of the $r$ distributions in Figure~\ref{fig:rDist}. Mean
locations for each distribution are indicated with vertical dashed lines; thano.
-0.0010; chrono. -0.0027. The distribution of thanatological $r$ is more compact
than for chronological $r$, with the ratio of variances (thano./chrono.) of
about 0.75. Usually the two theoretical values of $r$ move in the same direction, but thanatological $r$
is over time the less erratic of the two, and it usually paints a less dire picture when both are negative.

@@

The differences described here are not due primarily to rounding errors in the
data, but rather to the way in which $f^\star$ has been calculated. If the
births and exposure used to calculate $f^\star$ come from a population that is not stable
then $f^\star$ may be subject to some compositional ruses. Specifically, in this
case we have transformed fertility rates from empirical data with a particular
chronological age structure. If the chronological age structure was far from its
stable distribution, then it is hard to imagine that $f^\star$ would remain
fixed over time. Some simple contrived examples will be presented in a future
version of this paper in order to eliminate this problem.

From the projection matrix, the ratio of the largest to the second largest
eigenvalue, the damping ratio, is an indicator of the \textit{springiness} of
the stable population structure, $c_y$ or $c_a$, with respect to disturbances from the stable state
as determined by vital rate distributions. A higher damping ratio means that the
population structure oscillates back to its stable state faster, i.e., that
oscillations decrease in size more rapidly. The thanatological damping ratio was
greater than the Leslie damping ratio for all 1834 population-years included in
our empirical analysis. Leslie damping ratios ranged from 1.01258 to 1.0518,
while thanatological damping ratios ranged from 1.0455 to 1.0868. Again, this
was for the raw data, and I am not sure whether this consistent finding is a
natural property of the model or an artifact of the unstable starting state of
the data used to derive thanatological fertility rates. Intuitively, the
thanatological smoothing process should be an order of magnitude stronger than
the chronological smoothing process because each thanatological age class
(node) -- except for the oldest-old -- is connected directly to each other
thanatological age class in both directions. Again, I think contrived examples
will be necessary to shed more light on this finding.

>>>>

\section*{Some preliminary empirical findings}
The thanatological renewal equation and the discrete projection matrix can be
put to work with data. In this case a few options are available for determining
what kind of thanatological fertility rates to use. The two obvious choices are
to derive rates from the stable population implied by the chronological renewal
model, or to derive rates such as those from Figure~\ref{fig:FySpaghetti},
based on a particular stock of population.

Of the 1834 population-years of data on hand, we optimize $r$ from both
\eqref{eq:thanoren} and \eqref{eq:lotka}. In this sample, both versions of $r$
are only plausibly equal in a single instance. Usually thanatological $r$ is
greater than Lotka's $r$ (1373 cases). When Lotka's $r$ is positive (693 cases), thanatological $r$ is greater just over of 50\% of the time (356), but when Lotka's $r$ is negative,
thanatological $r$ is the greater of the two around 90\% of the time (1017
cases). These two approximations of $r$ are of opposite sign 138 times. We
provide a comparison of the $r$ distributions in Figure~\ref{fig:rDist}. Mean
locations for each distribution are indicated with vertical dashed lines; thano.
-0.0010; chrono. -0.0027. The distribution of thanatological $r$ is more compact
than for chronological $r$, with the ratio of variances (thano./chrono.) of
about 0.75. The two theoretical values of $r$ covary strongly, but
thanatological $r$ is the less erratic of the two, and it usually paints a less dire picture when both are negative.

@@

The differences described here are not due primarily to rounding errors in the
data, but rather to the way in which $f^\star(y)$ has been calculated. 
Specifically, in this case we have transformed fertility rates from empirical data with a particular
chronological age structure. If the chronological age structure is far from its
stable distribution, then it may be hard to imagine that $f^\star(y)$ will
remain fixed over time. When holding chronological-age classified fertility rates
fixed, the assumption is just the opposite.

From the projection matrix, the ratio of the largest to the second largest
eigenvalue, the damping ratio, is an indicator of the \textit{springiness} of
the stable population structure, $c^\star(y)$ or $c(a)$, with respect to
disturbances from the stable state as determined by vital rate distributions. A
higher damping ratio means that the population structure oscillates back to
its stable state faster, i.e., that oscillations decrease in size more rapidly. The thanatological damping ratio was
greater than the Leslie damping ratio for all 1834 population-years included in
our empirical analysis. Leslie damping ratios ranged from 1.01258 to 1.0518,
while thanatological damping ratios ranged from 1.0455 to 1.0868. Again, this
was for the raw data, and I am not sure whether this consistent finding is a
natural property of the model or an artifact of the unstable starting state of
the data used to derive thanatological fertility rates. Intuitively, the
thanatological smoothing process should be an order of magnitude stronger than
the chronological smoothing process because each thanatological age class
(node) -- except for the oldest-old -- is connected directly to each other
thanatological age class in both directions. Again, I think contrived examples
will be necessary to shed more light on this finding.
